---
title: "Data Manipulation"
output: github_document
date: "2022-09-22"
---

Once you've imported data, you're going to need to do some cleaning up.

```{r}
library(tidyverse)

options(tibble.print_min = 3) #print the first 3 rows

litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

## `select`
You can use the `select` function to specify the columns you want.

You can select which columns you want to keep by naming all of them:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

This tends to be tedious, so there are some shortcuts.


You can specify a range of variables to include:
```{r}
select(litters_data, group:gd_of_birth) 
# Select variables from group to gd_of_birth
```

If you need many variables, you can specify which columns you'd like to remove:
```{r}
select(litters_data, -pups_survive)
# Include all variables except for pups_survive (minus sign)
select(litters_data, -pups_survive, -group)
```


You can also use the `select` function to rename variables as you pull them:
```{r}
# new name on the left, original variable on the right of the equal sign
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

If you want to keep the original data set (no filtering) and just rename some variables, you can just use the `rename` function:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are many helper functions within `select` to make things easier and faster.
`starts_with()` to filter for specific variables starting with a character, so you don't have to list everything out:

```{r}
select(litters_data, starts_with("gd"))
```
The same goes for `ends_with()`

`everything()` can reorganize columns without discarding anything:
```{r}
# litter_number and pups_survive go first, then everything else
select(litters_data, litter_number, pups_survive, everything())
```

`relocate` does a similar thing (and is sort of like rename in that it’s handy but not critical):

```{r}
relocate(litters_data, litter_number, pups_survive)
```

## `filter`
You can use the `filter` function to specify rows instead of columns.
Typically use logical expressions to set criteria to select observations

```{r}
filter(pups_data, sex == 1)
filter(pups_data, sex == 2, pd_walk < 11)
```

You can use `drop_na()` within the function to remove missing variables

```{r}
drop_na(litters_data) # remove any row with a missing value
drop_na(litters_data, wt_increase) # remove rows for which wt_increase is missing.
```

## `mutate`

Use `mutate` to change or add new variables

```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group),
  wt_gain_kg = wt_gain * 2.2
)
```

## `arrange`

To rearrange rows in dataset, could be by a variable.

```{r}
head(arrange(litters_data, group, pups_born_alive), 10) # Sort groups, by ascending pups_born_alive
```

You can also go in ascending order:
```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10) # Sort descending groups, by ascending pups_born_alive

```


## `%>%`
AKA piping.
Pre-piping, you could need to create multiple iterations of your data set by step, very tedious!


```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

Another tedious way is to nest functions within functions, but you would have to kind of work backwards:
```{r}
litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )

litters_data_clean
```

The pipe is an easier alternative:

```{r}
# Use cmd+shift+m to call for the pipe
# Each step creates a tibble and passes it along to the next function
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% # load tibble
  janitor::clean_names() %>% # then clean
  select(-pups_survive) %>% # then select variables
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% # then edit variables
  drop_na(wt_gain) # then remove rows

```



