---
title: "Tidy Data"
output: github_document
date: "2022-09-27"
---

# `pivot_longer`

## Setup
Load packages and limit the number of rows to be displayed

```{r, message=F}
library(tidyverse)
options(tibble.print_min = 5)
```


## Load data

Data is a SAS file, need to use `haven` package.
Then clean names to be all lower case with clean_names.
```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()

pulse_df
```
The BDI score is spread across 4 columns (from bdi_score_bl to bdi_score_12m),
we want to use `pivot_longer` to bring them all to the same column.



```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", # make a column with the initial column names
    values_to = "bdi", # the values each column
    names_prefix = "bdi_score_") # removes "bdi_score_" from the visit column to only see what's remaining

pulse_tidy_data
```

We can also do everything in one step with the pipe.

```{r, eval=F}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi", 
    names_prefix = "bdi_score_") 

```

Let's mutate the "bl" label for visit into "00m"

```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi", 
    names_prefix = "bdi_score_") %>% 
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
                    visit = factor(visit)) %>% 
      arrange(id, visit)

pulse_df
```

## Learning assessment

In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?

```{r}
litters_df = read_csv("./data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  select(litter_number, gd0_weight, gd18_weight) %>% # can also use ends_with("weight")
  pivot_longer(gd0_weight:gd18_weight,
               names_to = "gd",
               values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))

litters_df
```

# `pivot_wider`

Sometimes untidy data is easier to understand than tidy data.
First, here is an example of tidy data:

```{r}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
)
analysis_result
```

Let's use `pivot_wider` to make it wider:
```{r}
pivot_wider(
  analysis_result, 
  names_from = "time",  # Column names coming from the "time" variable
  values_from = "mean") # What's in the dataframe from the "mean" variable
```

# `bind_rows`
Let's try it with the LOTR excel sheet that has 3 tables

```{r}
fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring") # add a column that show

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
```

Now that we laoded our separate datasets, let's bind rows together:

```{r}
lotr_bind = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names()
```

Better, but we could put the number of words for female and male columns together into one gender column:
```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  pivot_longer(
    female:male,
    names_to = "gender", 
    values_to = "words")   %>% 
  mutate(race = str_to_lower(race)) %>% # Change "Race" variable to lower case (race)
  select(movie, everything()) # Reorganize everything (no deletion) by the movie variable
```


# Joining datasets

First, let's load our separate datasets from pups and litters:
```{r}
pup_data = 
  read_csv("./data/FAS_pups.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = recode(sex, `1` = "male", `2` = "female"),
    sex = factor(sex)) 

litter_data = 
  read_csv("./data/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>% 
  # group is formatted as e.g. Con7 - separate it by 3 characters into Con and 7 in separate variables
  relocate(litter_number) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

Then we use a `left_join` to include all pups and all litter within pups (pups is the "left" data set)
```{r}
fas_data = 
  left_join(pup_data, litter_data, by = "litter_number")
```

`anti_join` would point out the data from the y dataset that could not be joined by the `left_join`













